---
title: "Project 2"
author: "Axel Paxel & Johneponne"
date: "8 maj 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

init

```{r}
library(ggplot2)

cdi <- read.delim("CDI.txt")
cdi$region <- factor(cdi$region, levels = c(1, 2, 3, 4),
                     labels = c("Northeast", "Midwest", "South", "West"))
cdi$crm1000 <- 1000 * cdi$crimes / cdi$popul


summary(cdi$crm1000)
median = 52.429

cdi <- cbind(cdi, hicrm = ifelse(cdi$crm1000 < median, 0, 1))

```


Plot hicrm against higrads, add a kernel smoother (see the lecture R-code for an example), and determine
visually whether there might be a relationship. Then fit a logistic regression using higrads as covariate and
add the estimated probabilities to the plot. Does it seem reasonable?


```{r}
with(cdi, {
   plot(hicrm ~ higrads)
   lines(ksmooth(higrads, hicrm, bandwidth = 20))
})

model <- glm(hicrm ~ higrads, data = cdi, family = "binomial")
summary(model)

x0 <- data.frame(higrads = seq(0, 100, 1))


predx <- cbind(x0, prob = predict(model, x0, type = "response"))
with(predx, lines(higrads, prob, col = "blue"))

# calculate conf.int for the linear part x*beta:
standard_error <- 1.96
xb <- predict(model, x0, se.fit = TRUE)
ci.xb <- data.frame(lwr = xb$fit - standard_error * xb$se.fit,
                    upr = xb$fit + standard_error * xb$se.fit)

# transform to CI for the odds:
ci.odds <- exp(ci.xb)

# and finally CI for the probabilities and add to the plot:
predx <- cbind(predx, ci.odds / (1 + ci.odds))
with(predx, {
  lines(higrads, lwr, lty = 2, col = "red")
  lines(higrads, upr, lty = 2, col = "red")
})


```
Seems to be relationship


Report the β-estimates together with their confidence intervals and test whether the amount of adults with
12 years in school has a significant effect on the probability of having a higher than median crime rate


```{r}
sum <- summary(model)
coeff.beta <- model$coefficients
coeff.beta_odds <- exp(coeff.beta) / (1 + exp(coeff.beta))
coeff.beta_odds

ci.beta <- confint(model)
ci.beta_odds <- exp(ci.beta) / (1 + exp(ci.beta))
```
Significance!

Estimate the relative change in the odds (odds ratio) of having a high crime rate, with confidence interval,
when the amount of higrads is increased by 1 (percent), and when it is increased by 10 (percent).

```{r}
exp(coeff.beta)["higrads"]
exp(coeff.beta)["higrads"]^10

increase_one_percent <- round((1 - exp(coeff.beta)["higrads"])*100, digits=1)
increase_ten_percent <- round((1 - exp(coeff.beta)["higrads"]^10)*100, digits=1)
```

If higrads increases 1%, odd increases by `r increase_one_percent`%
If higrads increases 10%, odd increases by `r increase_ten_percent`%


Use the model to predict the probability, with confidence interval, of having a high crime rate in a county
where the amount of higrads is 65 (percent), and where it is 85 (percent).
```{r}
x0 <- data.frame(higrads = c(65, 85))


predx <- cbind(x0, prob = predict(model, x0, type = "response"))

standard_error <- 1.96
xb <- predict(model, x0, se.fit = TRUE)
ci.xb <- data.frame(lwr = xb$fit - standard_error * xb$se.fit,
                    upr = xb$fit + standard_error * xb$se.fit)

# transform to CI for the odds:
ci.odds <- exp(ci.xb)

# and finally CI for the probabilities and add to the plot:
predx <- cbind(predx, ci.odds / (1 + ci.odds))

predx
```

Use the model to predict, for each of the counties, whether it would be expected to have a low or a high crime
rate (predicted probability below or above 0.5) and calculate the sensitivity and specificity1
for this model.

```{r}
predict_counties <- cbind(cdi, pred_crim = predict(model, cdi$higrads, type = "response"))
predict_counties

```

