---
title: "Predicting the burned area of a forest fire in northeastern Portugal"
author: "Axel Sjöberg & John Rapp Farnes"
date: "14 maj 2019"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
classoption: a4paper
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>"
)
options(tinytex.verbose = TRUE)
```

```{r setup_pdf, eval = FALSE}
# Install from CRAN
install.packages('rmarkdown')
# Render PDF
install.packages("tinytex")
tinytex::install_tinytex()  # install TinyTeX
```

```{r, code = readLines("lib.R")}
```

```{r, code = readLines("script.R")}
```


```{r}
titanic <- read.csv("titanic.csv")
titanic <- subset(titanic, select = -c(Name))
colnames(titanic) <- c(
  "survived",
  "pclass",
  "sex",
  "age",
  "siblings.spouses",
  "parents.children",
  "fare"
)
titanic$pclass <- factor(titanic$pclass, levels = c(1,2,3))

titanic
```

```{r}
splitted.dataset <- split.dataset(titanic)
(titanic.train <- splitted.dataset[[1]])
(titanic.test <- splitted.dataset[[2]])
```


<!-- Abstract? -->

# Introduction

<!-- Syfte -->

# Analysis

```{r}
ggpairs(titanic)
```

```{r}
hist(titanic$fare)
high.fare.passengers <- which(titanic$fare > 200)
low.fare.passengers <- which(titanic$fare <= 0)
titanic[high.fare.passengers,]
titanic[low.fare.passengers,]
length(high.fare.passengers)
length(low.fare.passengers)
```

```{r}
hist(titanic$parents.children)
which(titanic$parents.children > 2)
```


```{r}
# titanic <- titanic[-high.fare.passengers,]
#   
# hist(log(titanic$Fare))
titanic$log.fare <- log(titanic$fare + 1)
hist(titanic$log.fare)
```

```{r}
hist(log(titanic[which(titanic$pclass == 1),]$fare + 1))
hist(log(titanic[which(titanic$pclass == 2),]$fare + 1))
hist(log(titanic[which(titanic$pclass == 3),]$fare + 1))
```


```{r}
model.log.fare <- plot.binary.model(survived ~ log.fare, titanic, titanic$survived, titanic$log.fare, "log.fare", c(0, 500), 1)
plot.binary.model(survived ~ siblings.spouses, titanic, titanic$survived, titanic$siblings.spouses, "siblings.spouses", c(0, 10), 40)
plot.binary.model(survived ~ parents.children, titanic, titanic$survived, titanic$parents.children, "parents.children", c(0, 10), 40)
plot.binary.model(survived ~ age, titanic, titanic$survived, titanic$age, "age", c(0, 80), 40)
summary(model.log.fare)
dataset <- titanic
pR2(model.log.fare)["r2CU"]
```

```{r}
confint.beta <- calc.confint.beta(model.log.fare)

confint.OR <- calc.confint.OR(confint.beta)

cbind(beta = round(confint.beta[,"beta"], 1),
      OR = round(confint.OR, 1),
      "P-value" = round(confint.beta[,"P-value"], 4))

# create.table.png(confint.OR, "fare-OR")
```

```{r}
models <- list(
  glm(survived ~ fare, data = titanic, family = "binomial"),
  glm(survived ~ log.fare, data = titanic, family = "binomial")
)

model.names <- c(
  "fare",
  "log.fare"
)

compare.models <- calc.metrics.models(models, model.names)
# create.table.png(compare.models, "compare-fare-log")

cbind("AIC" = round(compare.models$aic, 0),
      "BIC" = round(compare.models$bic, 0),
      "Sensitivity (%)" = round(compare.models$sensitivity * 100, 1),
      "Specificity (%)" = round(compare.models$specificity * 100, 1),
      "Pseudo R^2 (%)" = round(compare.models$PseudoR2 * 100, 1))

```


## Survived vs age


Age vs survived
```{r}
model.age <- plot.binary.model(survived ~ age, titanic, titanic$survived, titanic$age, "age", c(0, 80), 40)
summary(model.age)
dataset <- titanic
pR2(model.age)["r2CU"]
```

Divide into age groups?
- Children saved first
- Mothers in certain age
- Seniors not prioritized

```{r}
titanic$age.group.even <- cut(titanic$age, seq(0, 80, 10))
cross.age.survive <- table(titanic$age.group.even, titanic$survived)
freq.table.age.survive <- prop.table(cross.age.survive, 1)
cbind(freq.table.age.survive, count = table(titanic$age.group.even))
plot(freq.table.age.survive[,2], xaxt = "n")
axis(1, at=1:nrow(freq.table.age.survive), rownames(freq.table.age.survive))
```

Good survival rate small children, then lower for young adults, higher for adults, low for seniors
Little data on 50+, merge into one

```{r}
```


```{r, fig.height=5, fig.width=10}
titanic$age.group <- cut(titanic$age, c(0, 15, 30, 55, 100), labels = c("Child",  "Young adult", "Adult", "Senior"))
titanic$age.group <- relevel(titanic$age.group, ref = "Adult")

cross.age.survive <- table(titanic$age.group, titanic$survived)
freq.table.age.survive <- prop.table(cross.age.survive, 1)
cbind(freq.table.age.survive, count = table(titanic$age.group))
plot(freq.table.age.survive[,2], xaxt = "n")
axis(1, at=1:nrow(freq.table.age.survive), rownames(freq.table.age.survive))
```

Quite spread out

```{r}
table(titanic$age.group, titanic$parents.children)
table(titanic$age.group, titanic$siblings.spouses)
```

Hard to distinquish children from parents and siblings from spouses

Next, fit model

```{r}
model.age.group <- glm(survived ~ age.group, data = titanic, family = "binomial")
(sum.age.group <- summary(model.age.group))
```

Significant

Look at beta, OR and prob with confidence interval

```{r}
confint.beta <- calc.confint.beta(model.age.group)

confint.OR <- calc.confint.OR(confint.beta)

cbind(beta = round(confint.beta[,"beta"], 1),
      OR = round(confint.OR, 1),
      "P-value" = round(confint.beta[,"P-value"], 4))

x0 <- data.frame(age.group = rownames(freq.table.age.survive))
calc.pred.prob(model.age.group, x0)
```


Comparing sole covariates

```{r}



sole.models <- list(
  glm(survived ~ parents.children, data = titanic, family = "binomial"),
  glm(survived ~ siblings.spouses, data = titanic, family = "binomial"),
  glm(survived ~ age.group, data = titanic, family = "binomial"),
  glm(survived ~ age, data = titanic, family = "binomial"),
  glm(survived ~ log.fare, data = titanic, family = "binomial"),
  glm(survived ~ pclass, data = titanic, family = "binomial"),
  glm(survived ~ is.alone, data = titanic, family = "binomial"),
  glm(survived ~ sex, data = titanic, family = "binomial")
)


sole.names <- c(
  "parents.children",
  "siblings.spouses",
  "age group",
  "age",
  "fare",
  "pclass",
  "is.alone",
  "sex"
)

(losmetricas <- calc.metrics.models(sole.models, sole.names))

plot.compare.models(losmetricas, c(900, 1250))

```









```{r}
models <- list(
  glm(survived ~ age, data = titanic, family = "binomial"),
  glm(survived ~ age.group, data = titanic, family = "binomial"),
  glm(survived ~ age.group.even, data = titanic, family = "binomial"),
  glm(survived ~ log.fare + age, data = titanic, family = "binomial"),
  glm(survived ~ log.fare + age.group, data = titanic, family = "binomial"),
  glm(survived ~ log.fare * age, data = titanic, family = "binomial"),
  glm(survived ~ log.fare * age.group, data = titanic, family = "binomial")
)

model.names <- c(
  "age",
  "age.group",
  "age.group.even",
  "log.fare + age",
  "log.fare + age.group",
  "log.fare * age",
  "log.fare * age.group"
)

compare.models <- calc.metrics.models(models, model.names)
# create.table.png(compare.models, "compare-fare-log")

cbind("AIC" = round(compare.models$aic, 0),
      "BIC" = round(compare.models$bic, 0),
      "Sensitivity (%)" = round(compare.models$sensitivity * 100, 1),
      "Specificity (%)" = round(compare.models$specificity * 100, 1),
      "Pseudo R^2 (%)" = round(compare.models$PseudoR2 * 100, 1))

lapply(models, function (model) summary(model)$coefficients)
```

```{r, fig.width = 12}
plot.compare.models(compare.models, c(1050, 1300))
```

Compare models, with fare


```{r}
models <- list(
  glm(survived ~ age, data = titanic, family = "binomial"),
  glm(survived ~ age.group, data = titanic, family = "binomial"),
  glm(survived ~ age.group.even, data = titanic, family = "binomial"),
  glm(survived ~ fare, data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1), data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) + age, data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) + age.group, data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) + age.group.even, data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) * age.group, data = titanic, family = "binomial")
)
model.names <- c(
  "age",
  "group",
  "even",
  "!log.fare",
  "fare",
  "fare + age",
  "fare + group",
  "fare + even",
  "fare * group"
)
compare.models <- calc.metrics.models(models, model.names)
compare.models
```

```{r, fig.width = 12}
plot.compare.models(compare.models, c(1050, 1250))
```

Remove the worst
```{r}
models <- list(
  glm(survived ~ log(fare + 1), data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) + age, data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) + age.group, data = titanic, family = "binomial"),
  glm(survived ~ log(fare + 1) * age.group, data = titanic, family = "binomial")
)
model.names <- c(
  "fare",
  "fare + age",
  "fare + group",
  "fare * group"
)
compare.models <- calc.metrics.models(models, model.names)
compare.models
```

```{r, fig.width = 8}
plot.compare.models(compare.models, c(1050, 1120))
```

## Sex

```{r}
boxplot(log.fare ~ sex, data = titanic)
boxplot(age ~ sex, data = titanic)
cross.sex.survive <- table(titanic$sex, titanic$survived)
cross.sex.survive

round(prop.table(cross.sex.survive, 1) * 100)
```

Much higher percentage of women survived. Most women survived while most men did not
No major differences in other factors

```{r}
titanic$sex <- relevel(titanic$sex, ref = "male")
model.sex <- glm(survived ~ sex, data = titanic, family = "binomial")
summary(model.sex)
```
```{r}
(confint.beta <- calc.confint.beta(model.sex))
(confint.OR <- exp(confint.beta))
x0 <- data.frame(sex = rownames(cross.sex.survive))
calc.pred.prob(model.sex, x0)
```

# Is alone

```{r}

titanic$is.alone <- ifelse(titanic$siblings.spouses + titanic$parents.children == 0, 1, 0)

table(titanic$is.alone)

(cross.table <- with(titanic, table(is.alone, survived)))
rownames(cross.table) <- c("Not alone", "Alone")
colnames(cross.table) <- c("Not survived", "Survived")

#create.table.png(cross.table, table.name = "is-alone_vs_survivors")

prop.table(cross.table, 1)
prop.table(cross.table, 2)
```

Most were alone

Of those who were alone, less survived

Of those who survived, about 50% were alone, of those who did not survive, most were alone


```{r}
model.is.alone <- glm(survived ~ is.alone, data = titanic, family = "binomial")
summary(model.is.alone)
```

```{r}
models <- list(
  glm(survived ~ parents.children, data = titanic, family = "binomial"),
  glm(survived ~ siblings.spouses, data = titanic, family = "binomial"),
  glm(survived ~ parents.children + siblings.spouses, data = titanic, family = "binomial"),
  glm(survived ~ is.alone, data = titanic, family = "binomial")
)

model.names <- c(
  "par/chil",
  "sib/spou",
  "par/chil+sib/spou",
  "alone"
)

compare.models <- calc.metrics.models(models, model.names)
# create.table.png(compare.models, "compare-fare-log")

cbind("AIC" = round(compare.models$aic, 0),
      "BIC" = round(compare.models$bic, 0),
      "Sensitivity (%)" = round(compare.models$sensitivity * 100, 1),
      "Specificity (%)" = round(compare.models$specificity * 100, 1),
      "Pseudo R^2 (%)" = round(compare.models$PseudoR2 * 100, 1))

```

```{r, fig.width = 12}
plot.compare.models(compare.models, c(1125, 1200))
```



# Pclass
```{r}

cross.pclass.survive <- table(titanic$pclass, titanic$survived)
cross.pclass.survive

round(prop.table(cross.pclass.survive, 1) * 100)

```

```{r}
model.pclass <- glm(survived ~ pclass, data = titanic, family = "binomial")
summary(model.pclass)
```

```{r}
models <- list(
  glm(survived ~ log.fare, data = titanic, family = "binomial"),
  glm(survived ~ pclass, data = titanic, family = "binomial")
)

model.names <- c(
  "log.fare",
  "pclass"
)

compare.models <- calc.metrics.models(models, model.names)
# create.table.png(compare.models, "compare-fare-log")

cbind("AIC" = round(compare.models$aic, 0),
      "BIC" = round(compare.models$bic, 0),
      "Sensitivity (%)" = round(compare.models$sensitivity * 100, 1),
      "Specificity (%)" = round(compare.models$specificity * 100, 1),
      "Pseudo R^2 (%)" = round(compare.models$PseudoR2 * 100, 1))

```

```{r, fig.width = 12}
plot.compare.models(compare.models, c(1070, 1120))
```

# Compare models


```{r}
models <- list(
  glm(survived ~ sex, data = titanic, family = "binomial"),
  glm(survived ~ sex + age + pclass + log.fare + siblings.spouses + parents.children, data = titanic, family = "binomial"),
  glm(survived ~ log.fare * age * sex * pclass * is.alone, data = titanic, family = "binomial"),
  glm(survived ~ log.fare * age * sex * pclass * is.alone * parents.children + siblings.spouses, data = titanic, family = "binomial"),
  glm(survived ~ log.fare * age * sex * pclass * is.alone + parents.children + siblings.spouses, data = titanic, family = "binomial"),
  glm(survived ~ pclass + age * sex + is.alone, data = titanic, family = "binomial"),
  glm(survived ~ pclass * age * sex * is.alone, data = titanic, family = "binomial")
)
titanic
model.names <- c(
  "s",
  "s+a+p+f+s+p.c",
  "interaction some 1",
  "interaction some 2",
  "interaction some 3",
  "John 1",
  "John 2"
)
(compare.models <- calc.metrics.models(models, model.names))

# models[[6]]
# summary(models[[6]])
```











```{r, fig.width = 12}
plot.compare.models(compare.models, c(750, 1150))
```

<!-- A/B -->
```{r}
models <- list(
  glm(survived ~ log.fare + age, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age.group, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age + sex, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age.group + sex, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age + sex + pclass, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age + sex + pclass + siblings.spouses + parents.children, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age + sex + pclass + is.alone, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare * age * sex * pclass * siblings.spouses * parents.children, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare * age * sex * pclass * is.alone, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age * parents.children * siblings.spouses + sex + pclass, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age + sex + pclass + siblings.spouses, data = titanic.train, family = "binomial"),
  glm(survived ~ log.fare + age.group + sex + pclass + siblings.spouses, data = titanic.train, family = "binomial")
)
titanic
model.names <- c(
  "f+a",
  "f+ag",
  "f+a+s",
  "f+ag+s",
  "f+ag+s+c",
  "all",
  "all+alone",
  "all interaction",
  "all+alone interaction",
  "some interaction",
  "no parents",
  "no parents, group"
)
compare.models <- calc.metrics.models(models, model.names)
compare.models
# models[[6]]
# summary(models[[6]])
```
<!-- A/B -->


The three top performing models as can be seen in the digram above are: 

All,

All interaction,

Some, interaction


Down below, a comparison of the three models unfolds dadada.



QQ-plot

```{r}
model.all <- glm(survived ~ log.fare + age + sex + pclass + siblings.spouses + parents.children, data = titanic, family = "binomial")

model.all.interaction <- glm(survived ~ log.fare * age * sex * pclass * siblings.spouses * parents.children, data = titanic, family = "binomial")

model.some.interaction <- glm(survived ~ log.fare + age * parents.children * siblings.spouses + sex + pclass, data = titanic, family = "binomial")

best.models <- list(
  model.all,
  model.all.interaction,
  model.some.interaction
)

lapply(best.models, qq)




```

model.interaction.some seem to be in best allignment with a normal distribution. The others dont fall to short though. In order to value the influence of outliers, Cook's distance plots for the three models are plotted below. 


Cook's

```{r}
cook.plotter(model.all, titanic)

sapply(best.models, cook.plotter, titanic)

```

DF-betas

```{r}
# dfbeta.plot(model.all)
# dfbeta.plot(model.all.interaction)
# dfbeta.plot(model.some.interaction)

```

From the DFbeta plots it looks like the all.interaction has several issues. For the DFbeta.parents.children there are several points that are with quite the marignal above the horizontal lines. For the all model and the some.interaction model, The points in the plots are a lot more centered around the x axis, and the outliers are not as far away, (Fewer and closer). In the all.interaction model we want to look at the data entires that causes DF-beta values above 0.5. 



```{r}
dfb <- dfbetas(model.all.interaction)
#dfb
d.Intercept <- as.vector(dfb[,1])
d.log.fare <- as.vector(dfb[,2])
d.age <- as.vector(dfb[,3])
d.sexfemale <- as.vector(dfb[,4])
d.pclass <- as.vector(dfb[,5])
d.siblings.spouses <- as.vector(dfb[,6])
d.parents.children <- as.vector(dfb[,7])
d.log.fare

outlier.index1 <- which(d.Intercept > 0.5) #372
(outlier.index2 <- which(d.log.fare > 0.5)) 
(outlier.index3 <- which(d.age > 0.5))
(outlier.index4 <- which(d.sexfemale > 0.5))
(outlier.index5 <- which(d.pclass > 0.5))
(outlier.index6 <- which(d.siblings.spouses > 0.5)) #503 
(outlier.index7 <- which(d.parents.children > 0.5)) #97 118 677  (where    677 > 1) 





```
Entry 372 was a 22 year old man, alone, payed a substantial amount and stayed in first class. 

Entry 503 was an 18 year old man, had one sibling or spouse on board, payed a substantial amount and stayed in first class

entry 97 was a 23 year old man with no spouse and either a parent or a child onboard. The man survived. He stayed in first class and payed alot more than the average. Entry 118 was a 24 year old man with no spouse and either a parent or a child onboard. This man also stayed in first class and payed a substantial amount. Entry 677 was a 36 year old man that payed a substantial amount (highest of all the passengers), stayed in first class. He had one parent or child on board and 
















DOWN MIGHT BE WRONG.
The optimal model acording to us is the the some interaction model. While it still enjoys a rather large Pseudo R2 it furthermore boasts a low AIC and BIC. There was one model that enjoyed a higher Pseudo R2 than the decided optimal model, however it had plenty of interaction terms which made it more complex. Larger models tend to perform worse on the BIC. The relative increse in Pseudo R2 does not in this case compensagte for the increased complexity in the model, which can be seen in the Spike of the digaram. From the optimal, a backward selection is performed:


Backward Selection:
```{r}
optimal.model <- glm(survived ~ log.fare + age * parents.children * siblings.spouses + sex + pclass, data = titanic, family = "binomial")
calc.metrics(optimal.model)

```

No change in removing any varible caused the Pseudo R2 to drop compared to the original model. Thus the optimal model reminas unchanged.  

Story:
- Dataset, kvalitativt. Berätta lite om titanic
- Syfte: hitta så bra modell som möjligt mätt på de här metrics. Med den göra predictions för de här personerna
- Dataset, columner mm vad de betyder, tekniskt
- Feature engineering
  - Gå igenom alla features
  - Lite plots, fits, modeller
  - Hypoteser, försöka förklara resultat
- Modeller och comparison
  - Alla metrics, i.e AIC/BIC etc
  - Utesluta modeller
  - 2-3 key modeller, kolla etc
- Optimala modellen
  - Metrics
  - Förklaring till varför den funkar
  - Cooks etc
- Resultat
  - Metrics
  - A/B
  - Slide med de här personerna och deras survival rates

John:
- Is alone analys
- Lägga till sense och spec i grafer
- Fortsätta på deck
- Dela upp i A/B
Axel
- Dataset, Syfte, Dataset (utkast)
- Börja på feature engineering